---
title: Integrative Analysis of Multi-omic Data
author:
  - name: Piero Palacios Bernuy
    orcid: 0000-0001-6729-4080
    corresponding: true
    email: p.palacios.bernuy@gmail.com
    roles:
      - Investigation
      - Bioinformatics
      - Deep learning
      - Visualization
keywords:
  - Genomic Ranges
  - Omics
  - Bioconductor
abstract: |
  This document is part of a series of the analysis of Omics data. Especifically, here is showed how to analyze bulk RNA-Seq data with Bioconductor packages. Also, it's showcased how to make plots of the RNA data in the context of differentially gene expression and gene-sets. 
plain-language-summary: |
  This document have a example of the analysis of bulk RNA-Seq data.
key-points:
  - A guide to analyze GWAS public data.
  - A guide to analyze TCGA database.
date: last-modified
bibliography: references.bib
citation:
  container-title: An open source portfolio
number-sections: true
execute: 
  echo: true
---

## Introduction



## GWAS Catalog with a ChIP-Seq Experiment

Here,  we are gonna analyze the relation between transcription factor binding (ESRRA binding data) from a ChIP-Seq experiment and the genome-wide associations between DNA variants and phenotypes like diseases. For this task, we are gonna use a the `gwascat` package distributed by the **EMBL** (European Molecular Biology Laboratories).


```{r}
library(tidyverse)
library(gwascat)
library(GenomeInfoDb)
library(ERBS)
library(liftOver)
```

First, we need to download the data, keep the 24 chromosomes (from 1 to Y) and, specify the sequence information from the GRCh38 human genome annotation.

```{r}

gwcat = get_cached_gwascat()

gg = gwcat |> as_GRanges()

gg = keepStandardChromosomes(gg, pruning.mode = "coarse")

# seqlevelsStyle(gg) <- "UCSC"

seqlevels(gg) <- seqlevels(gg) |> 
  sortSeqlevels()

data("si.hs.38")

seqinfo(gg) <- si.hs.38

```

Now, let's plot a karyogram that will show the SNP's identified with significant associations with a phenotype. The SNP's in the GWAS catalog have a stringent criterion of significance and there has been a replication of the finding from a independent population.

```{r}
ggbio::autoplot(gg, layout="karyogram")
```

We can see the peak data as a `GRanges` object: 

```{r}
data("GM12878")

GM12878

```

If we see the bootom of the GRanges table, this experiment have the hg19 annotation from the human genome. To work on the GRCh38 annotation we need to liftover with a `.chain` file. For this we can use the `AnnotationHub` package.

```{r}
library(AnnotationHub)
ah <- AnnotationHub::AnnotationHub()

query(ah, c("hg19ToHg38.over.chain"))

chain <- ah[["AH14150"]]

```

```{r}
GM12878 <- liftOver(GM12878, chain) |> 
  unlist()

seqlevelsStyle(GM12878) <- "NCBI"

seqinfo(GM12878) <- si.hs.38


seqlevelsStyle(GM12878) <- "UCSC"
seqlevelsStyle(gg) <- "UCSC"
```

Now lets find overlaps between the GWAS catalog and the ChIP-Seq experiment.

```{r}
# duplicated loci
length(gg) - length(reduce(gg))

#
fo = findOverlaps(GM12878, reduce(gg))
fo


```


```{r}
over_ranges <- reduce(gg)[subjectHits(fo)]


ii <- over_ranges |> 
  as.data.frame() |> 
  GenomicRanges::makeGRangesListFromDataFrame()


phset = lapply(ii, function(x){
  
  # print(glue::glue("On range: {x}"))
  unique(gg[which(gg %over% x)]$"DISEASE/TRAIT")
  
})

gwas_on_peaks <- phset |> 
    enframe() |> 
    unnest(value) |> 
    count(value) |> 
    slice_max(n, n=20)


p <- gwas_on_peaks |> 
  mutate(value = fct_reorder(value,n)) |> 
  ggplot(aes(n,value, fill=value))+
  geom_col() +
  theme(legend.position = "none") +
  paletteer::scale_fill_paletteer_d("khroma::smoothrainbow") +
  theme_minimal()

bslib::card(plotly::ggplotly(p), full_screen = T)
```

distinct phenotypes identified on the peaks

```{r}
length(phset)
```
how to do the inference of these phenotype on peaks of these b cells

```{r}
library(ph525x)
library(progress)

set.seed(123)

n_iter = 100

pb <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed time: :elapsedfull || Estimated time remaining: :eta]",
                       total = n_iter,
                       complete = "=",   # Completion bar character
                       incomplete = " ", # Incomplete bar character
                       current = ">",    # Current bar character
                       clear = FALSE,    # If TRUE, clears the bar when finish
                       width = 100)   


rsc = sapply(1:n_iter, function(x) {
    pb$tick()
    length(findOverlaps(reposition(GM12878), reduce(gg))) |> 
    suppressWarnings()
    
})

# compute prop with more overlaps than in observed data
mean(rsc > length(fo))

```



## Data & Methods {#sec-data-methods}

{{< embed notebooks/TCGA.qmd#fig-genesmut >}}


## Conclusion

## References {.unnumbered}

:::{#refs}

:::