---
title: "Integrative Analysis with TCGA Data"
subtitle: "Analysis of Mutation Data from The Cancer Genome Atlas (TCGA)"
format: 
  html:
    code-fold: show
    smooth-scroll: true
    theme: minty
    toc: true
    toc-depth: 10
editor: visual
---

```{r}
#| include: false
library(knitr)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 180,
  fig.width = 15,
  fig.height = 8,
  echo = TRUE
)

library(tidyverse)

theme_set(theme_minimal())
```

## Intro


```{r}
library(curatedTCGAData)
library(TCGAutils)
library(MultiAssayExperiment)

```


Download the data

A complete list of disease codes used by the TCGA project are available on the [NCI Genomic Data Commons website](https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations).

```{r}

readData = curatedTCGAData("READ", 
                           c("RNASeq2GeneNorm", "Mutation", "Methylation_methyl450"), 
                           dry.run = FALSE, version = "2.1.1")

readData

```
```{r}
sampleMap(readData)
```
```{r}
sampleMap(readData) |> 
  as_tibble() |> 
  pull(primary) |> 
  table() |> 
  table()
```
```{r}
clin = colData(readData) |> 
  as_tibble()
dim(clin)

head(colnames(clin), 10) 

```

```{r}

clin |> 
  pull(pathology_T_stage) |> 
  table()
  
```
```{r}

clin <- clin |> 
  mutate(t_stage = case_when(
    pathology_T_stage %in% c("t4","t4a","t4b") ~ "t4",
    .default = pathology_T_stage
  ))

clin$t_stage |> 
  table()

```

```{r}
clin$vital_status |> 
  table()
```
```{r}
table(clin$t_stage, clin$vital_status)
```

# Analyzing Mutation Data

```{r}
mut_data = readData[[1]]

mut_data
```
```{r}
mut_sample_ids = colnames(mut_data)
head(mut_sample_ids)

head(clin$patientID)

```
```{r}

mut_sample_ids <- mut_sample_ids |> 
  stringr::str_sub(1,12)

all(mut_sample_ids %in% clin$patientID)

```

```{r}
assay(mut_data)[1:3,1:3]

# Is a sparse matrix

assay(mut_data)[1,] |> 
  table(useNA="ifany")

mut_assay = mut_data@assays

mut_assay

mut_assay |> class()

length(mut_assay)

```


```{r}
mut_assay[[1]]

mut_assay[[1]]$Hugo_Symbol

mut_assay[[1]]$Mutation_Status |> 
  table()

mut_assay[[1]]$Variant_Classification |> 
  table()

```


lets get all mutation information for all the samples

```{r}

var_class_df = mapply(function(sample_id, mutation_assay){
  
  d = mcols(mutation_assay)[,c("Hugo_Symbol","Variant_Classification")] |> 
    as.data.frame()
  
  colnames(d) = c("symbol","variant_class")
  
  d$patientID = sample_id
  
  return(d)
  
}, sample_id=mut_sample_ids, mutation_assay = mut_assay,SIMPLIFY = F, USE.NAMES = F)


var_class_df = do.call(rbind, var_class_df)


head(var_class_df)

```

most common mutated genes genes in rectum adenocarcinoma

```{r}
#| label: fig-genesmut

p <- var_class_df |> 
  as_tibble() |> 
  group_by(symbol,variant_class) |> 
  summarise(n = n()) |> 
  arrange(desc(n)) |> 
  ungroup() |> 
  slice_max(order_by = n, n = 20) |> 
  ungroup() |> 
  mutate(symbol = fct_reorder(symbol,n)) |> 
  ggplot(aes(n, symbol,fill=variant_class)) +
  geom_col() +
  facet_wrap(~variant_class) +
  paletteer::scale_fill_paletteer_d("awtools::a_palette") +
  labs(x = "Number of samples with specific mutation", y="Gene Symbol", fill="Variant Class",
       title="Top 20 Samples with Rectum Adenocarcinoma Mutation")


p

```


# Linking mutations and tumor stage

```{r}
index <- which( (var_class_df$patientID |> unique()) %in% clin$patientID)

clin_and_mutation = clin[index,]

head(clin_and_mutation) 

```

```{r}
#| label: fig-plot1


df = var_class_df |> 
  group_by(patientID) |> 
  summarise(n = n())

p <- clin_and_mutation |> 
  left_join(df, by = join_by(patientID)) |> 
  ggplot(aes(t_stage,log(n), fill=t_stage))+
  geom_boxplot() +
  geom_jitter(width = 0.05, colour="red2") +
  paletteer::scale_fill_paletteer_d("awtools::a_palette")
p
```

```{r}
#| label: fig-plot_tp53


mut_on_tp53 <- var_class_df |> 
  dplyr::filter(symbol == "TP53") 


p <-clin_and_mutation |> 
  dplyr::filter(patientID %in%  mut_on_tp53$patientID) |> 
  dplyr::select(t_stage, patientID) |> 
  group_by(t_stage) |> 
  summarise(n = n()) |> 
  # mutate(t_stage = fct_reorder(t_stage,n)) |> 
  ggplot(aes(n, t_stage, fill=t_stage, label=n)) +
  geom_col() +
  geom_label(fill="white")  +
  paletteer::scale_fill_paletteer_d("awtools::a_palette")+
  labs(y = "Tumor Stage", x="N° of Mutations", fill="Tumor Stage", title="Number of Mutations on Gene TP53")

p
```


# Linking expression and tumor stage


```{r}
rnaseq = readData[[2]]

rnaseq

```
```{r}

colnames(rnaseq) = colnames(rnaseq) |> 
  stringr::str_sub(1,12)

clin <- clin |> as.data.frame()
rownames(clin) <- clin$patientID

index <- which(colnames(rnaseq) %in% rownames(clin))

colData(rnaseq) = as(clin[colnames(rnaseq),],"DataFrame")


idx <- is.na(colData(rnaseq)$t_stage)

rnaseq <- rnaseq[,!idx]

```


```{r}

library(limma)
mm = model.matrix(~t_stage, data=colData(rnaseq))
f1 = lmFit(assay(rnaseq), mm)
ef1 = eBayes(f1)
topTable(ef1) |> 
  arrange(desc(AveExpr))

```


```{r}
par(mfrow=c(1,2))
boxplot(split(assay(rnaseq)["PAM",], rnaseq$t_stage), main="PAM")    # higher expression in lower t_stage
boxplot(split(assay(rnaseq)["PAIP2",], rnaseq$t_stage), main="PAIP2")
```



# Linking methylation and expression


```{r}

methyl <- readData[[3]]

idx <- colnames(methyl) |> 
  str_split(pattern = "-") |> 
  map(4) |> 
  enframe() |> 
  unnest(value)

idx <- which(idx$value == "01A")

methyl = methyl[,idx]

methyl
```
```{r}
colnames(methyl) <- colnames(methyl) |> 
  str_sub(start = 1,end = 12)
```

```{r}
colData(methyl) <- as(clin[colnames(methyl),],"DataFrame")


intersect(colnames(methyl), colnames(rnaseq)) |> length()

```

```{r}

methyl_subset = methyl[,which(colnames(methyl) %in% colnames(rnaseq))]
rnaseq_subset = rnaseq[,which(colnames(rnaseq) %in% colnames(methyl))]

methyl_genes = rowData(methyl_subset)
methyl_genes

```

```{r}

#| label: fig-plot

me_rna_cor = function(sym, mpick = 3){
    require(GGally)
    # subset methylation data to first mpick methylation sites for given gene symbol
    methyl_ind = which(methyl_genes$Gene_Symbol == sym)
    if (length(methyl_ind) > mpick){    
        methyl_ind = methyl_ind[1:mpick]
    }
    methyl_dat = assay(methyl_subset)[methyl_ind,]    # subset to selected methylation sites

    # subset expression data to selected gene symbol
    expr_ind = which(rownames(rnaseq_subset) == sym)    
    expr_dat = assay(rnaseq_subset)[expr_ind,]

    # combine methylation and expression data as data frame
    combined_dat = as(t(methyl_dat), "DataFrame")
    combined_dat$expr = expr_dat

    # plot pairs and calculate correlation coefficients between methylation and expression
    ggpairs(combined_dat) |> print()
    sapply(1:mpick, function(i){
      
        cor_to <- as.numeric(combined_dat[,mpick+1])
        
        df <- data.frame(v1= as.numeric(combined_dat[,i]),
                         v2 = cor_to)
        
        df <- df |> na.omit()
      
        cor(df[,1], df[,2])
    })
}

me_rna_cor("TAC1", mpick=3)

```








